---
title: "lab3yg"
format:
 html:
   embed-resources: true
---
Question1: 
```{r}
met <- read.csv("~/Downloads/met_all.gz")
```




Question 2: 
```{r}
dim(met)
head(met)
tail(met)
```

there are 2377343 rows and 30 column. 

Question 3:

```{r}
str(met)
```

Question4: 
```{r}
table(met$year)
```


```{r}
table(met$day)
```

```{r}
table(met$hour)
```

```{r}
summary(met$temp)
```

```{r}
summary(met$elev)
```

```{r}
summary(met$wind.sp)
```


```{r}
met$elev[met$elev==9999.0] <- NA

summary(met$elev)
```

The highest weather station is 4113 meters. 

```{r}
met <- met[met$temp>-40,]
met2 <- met[order(met$temp),]
head(met2)
```

Question 5

```{r}
met$elev[which.min(met$temp)]
range(met$elev,na.rm = TRUE)
```

```{r}
met <- met[met$temp>-15,]
met2 <- met[order(met$temp),]
head(met2)
```

Question 6:
```{r}
elev <- met[met$elev==max(met$elev, na.rm=TRUE), ]
summary(elev)
```


```{r}
cor(elev$temp, elev$wind.sp, use="complete")
cor(elev$temp, elev$hour, use="complete")
cor(elev$wind.sp, elev$day, use="complete")
cor(elev$wind.sp, elev$hour, use="complete")
cor(elev$temp, elev$day, use="complete")
```

Question 7;

```{r}
hist(met$elev, breaks=100)
hist(met$temp)
hist(met$wind.sp)
```
library(leaflet)
```{r}
library(leaflet)
leaflet(elev) %>%
  addProviderTiles('OpenStreetMap') %>% 
  addCircles(lat=~lat,lng=~lon, opacity=1, fillOpacity=1, radius=100)
```

```{r}
library(lubridate)
elev$date <- with(elev, ymd_h(paste(year, month, day, hour, sep= ' ')))
summary(elev$date)
elev <- elev[order(elev$date),]
head(elev)
```

Question 8:
```{r}
plot(elev$date, elev$temp, type='l')
plot(elev$date, elev$wind.sp, type='l')
```


Tem: -axis (horizontal): Dates ranging from early August to early September (Aug 01 – Sep 02). Y-axis (vertical): Temperature values (elev$temp), ranging from about 2 to 15. Temperatures oscillate daily (short-term cycles), likely reflecting day–night variation.Values generally fall between 5–12°C, with occasional lows around 2–4°C and highs peaking at 13–15°C.
Early August (Aug 01–10): Strong fluctuations, but temperatures mostly in mid-range (6–12°C).
Mid-August (Aug 11–20): Slight dips observed, some lows near 2–4°C.
Late August (Aug 21–31): More pronounced variability, with frequent swings between 5–13°C.

Wind: X-axis (horizontal): Dates ranging from early August to early September (Aug 01 – Sep 02).Y-axis (vertical): Wind speed values (elev$wind.sp), ranging approximately from 0 to a little over 20.  Wind speed fluctuates frequently with short-term variability. Early August (Aug 01–10): Wind speed is mostly lower, with values generally between 0–10. Mid-August (Aug 11–20): Increased variability, with some peaks approaching 15–20. Late August (Aug 21–30): More frequent high wind speeds, often above 10 and peaking above 20.


Question 8: 
```{r}


```


